The Soulgem Paper

### 2011-09-22 soulgem.tex
Nearly done with first draft of Findings section --just one and a half subsections left.

I'm leaving the abstract until later.

\end{abstract}

@@

\subsection{Dysfunction and role fluctuation}

@@

\item Performance that diverges from expectations, either positively or negatively, affects future expectations. Therefore, past performance in an organization causes \emph{fluctuations} in its role definitions.
\end{itemize}

@@

A good analysis of role fluctuation comes from Nicholson \cite{Nicholson1984}. He studied how people adapt to new work roles, and determined there are four main ways in which they do so, based on how much they change themselves and the environment: replication (neither the person nor the environment change much), absorption (personal development), determination (role development), and exploration (the person develops both herself and the meaning of her role).

@@

We can also find rich descriptions of some roles in software development. Martin \emph{et al.}\ \cite{Martin2007} report a detailed picture of the testing role carried out by the developers in a software organization, and of the reasons for which they deviate from the testing prescriptions in the literature. F{\ae}gri \emph{et al.} \cite{Faegri2010} provide a rich report of the consequences of an attempted role fluctuation in a case in which developers attempted to rotate the job of customer support.

@@

\textbf{\emph{RQ5:}} Why are some expectations consistently unsatisfied, and what consequences does such dissatisfaction bring?

\emph{Proposition:} Expectations may be consistently unsatisfied due to (a) poor performance, (b) unreasonable demands, (c) role strain, or (d) coordination problems.

\textbf{\emph{RQ6:}} Are there purely technical roles? Can we disassociate organizational and social concerns in technical role descriptions or prescriptions?

\emph{Proposition:} There are no purely technical roles. There are organizational and social concerns intertwined with the technical expectations placed on all members of the organization.

\textbf{\emph{RQ7:}} How do roles change over time in a software organization?

\emph{Proposition:} There are two main drivers of role fluctuation: organizational growth, which leads to specialization \cite{Blau1971,Haveman1993}, and organizational dysfunction, which leads to rearrangements of expectations.

\textbf{\emph{RQ8:}} What media do people enacting similar roles use to communicate to each other? What media do they use to communicate with others?

@@

% There is a confusion between roles and positions: We have positions that attempt to approximate roles, but they're not the same thing. A role is a set of responsibilities that people place on a person by virtue of his or her background, group membership, experience, position in the organization, cultural expectations, and stated commitments.

% The fact that expectations are placed does not mean that they are satisfied. (The expectee may not know or understand the expectation; the expectation may be beyond her means; the expectation may cause a role stress, etc.)

% Therefore a dynamic of expectations that works in theory may not work in practice. Similarly, a dynamic that *doesn't* work in theory may work in practice, due to the initiative of some people who contribute beyond what the rest expects of them. However, in both cases, over time the organization grows to expect the typical performance of others (of the under/over performers), and a new understanding (a new expectation,  a new norm) takes hold.

% Roles, even technical roles, include a number of organizational and social expectations that its enactors should satisfy. No role is purely technical, and often the organizational / social / political is a prime concern.

% On growth

% As an organization grows, labour is divided further, roles become more specialized. Some expectations that took a fraction of someone's time now are more demanding, hence the tendency to formalize them as an important part of a new role. But note that organization members can fight the specialization trend.

% Some expectations arise or are solidified due to the weight of the organization or to deal with some sort of dysfunction or structural situation.

@@

% On the expectations that occur over and over (move to earlier?)

% In a software organization, some expectations arise over and over. We can consider them a necessity for these organizations. A list will be arbitrary perhaps (we will almost surely miss some important ones?) but listing some key expectations is still useful.
% - Develop production code
% - Explore / scout development
% - Develop configuration code
% - Develop low level (unit) tests
% - Develop high level (integration) tests
% - Assign tasks
% - Review progress
% - Set direction, vision goals
% - Resolve interpersonal conflicts
% - Represent users
% - Take care of what falls between the cracks
% - and so on...
% I think it will be worth listing all the important ones. Perhaps more than a list? The key ideas here are that (a) these expectations *need* to be satisfied, or the performance of the organization will suffer, (b) some of them are conceptually closer to each other, and having them be carried out by the same person or persons should be more effective from a coordination perspective (see Proportionality, Maturity), and (c) understanding these patterns of responsibility pairing should yield efficiency gains more generalized.

>>>>

Let me not to the marriage of true minds / Admit impediments, love is not love / Which alters when it alteration finds, / Or bends with the remover to remove.

O no, it is an ever fixed mark / That looks on tempests and is never shaken; / It is the star to every wand'ring bark, / Whose worth's unknown although his height be taken.

Love's not time's fool, though rosy lips and cheeks / Within his bending sickle's compass come, / Love alters not with his brief hours and weeks, / But bears it out even to the edge of doom:

If this be error and upon me proved, / I never writ, nor no man ever loved.
\end{abstract}

@@

\subsection{Dysfunction and role evolution}

@@

\item Performance that diverges from expectations, either positively or negatively, affects future expectations. Therefore, past performance in an organization causes \emph{evolutions} in its role definitions.
\end{itemize}

@@

A good analysis of role evolution comes from Nicholson \cite{Nicholson1984}. He studied how people adapt to new work roles, and determined there are four main ways in which they do so, based on how much they change themselves and the environment: replication (neither the person nor the environment change much), absorption (personal development), determination (role development), and exploration (the person develops both herself and the meaning of her role).

@@

We can also find rich descriptions of some roles in software development. Martin \emph{et al.}\ \cite{Martin2007} report a detailed picture of the testing role carried out by the developers in a software organization, and of the reasons for which they deviate from the testing prescriptions in the literature. F{\ae}gri \emph{et al.} \cite{Faegri2010} provide a rich report of the consequences of an attempted role evolution in a case in which developers attempted to rotate the job of customer support.

@@

\textbf{\emph{RQ5:}} How do roles change over time in a software organization?

\emph{Proposition:} There are two main drivers of role evolution: organizational growth, which leads to specialization \cite{Blau1971,Haveman1993}, and organizational dysfunction, which leads to rearrangements of expectations.

\textbf{\emph{RQ6:}} What media do people enacting similar roles use to communicate to each other? What media do they use to communicate with others?

@@

As Table X shows, we found a considerable divergence between people's positions and the sets of expectations placed on them---their roles. However, there is an overlap: one can reasonably expect a Developer to write code of some kind, a QA Analyst to be concerned with software quality in one way or another, and a Team Lead to monitor and report on the team's progress. Table X provides an abstracted view of this finding, which also holds throughout our more detailed qualitative data: while there was no full agreement on expectations between people holding similar positions, some patterns and tendencies become apparent. We believe that this partial agreement comes from the fact that a person's position is, first, a statement about her placement in the organizational structure, and second, a partial statement of the institutional expectations placed on her. That is, those in the organization with the requisite authority give her a label that approximates their expectations (though perhaps not those of the rest of the organization) and place her in a position in the structure that should help her satisfy them.

However, a person's position has limited power in defining the person's role. As we have seen above, the expectations placed on a person also come from historical and personal factors. An example out of several in our observations comes from one of the Gallium Product Managers. He was hired as a Product Manager to steer and coordinate the development of the company's products, but he had no experience in software development, thought that his main asset was his understanding of and his contacts in the products' domain, so that he was in the organization as the ``voice of the customer.'' In his early days in the organization he struggled to understand the level of control and direction he was expected to have. Over time he learned to fill the void of control that he was expected to fill, but he continued to make use of his background, connections, and domain intuition to give his role a distinctly different flavour than those of his peers with the same or similar positions.

\subsection{Role Evolution (RQ5)}

We proposed that roles evolve for two main reasons: organizational growth and dysfunction. Our observations provide support for both reasons, but add a qualification to the second: it is not just dysfunction that drives role evolution, but any mismatches, positive or negative, between expectations and behaviours.

Regarding organizational growth, there is a long sociological tradition exploring its role in specialization and division of labour, and our data are far from surprising. In fact, it should be clear from pure argument that organizational growth drives role evolution. The smallest organization is formed by a single person, and that person should perform all the activities required for the functioning of the organization, from sales to debugging. As the organization gets more members, labour begins to be divided, not by amount, but by kind, so that upon reaching even a handful of members it is rare for a software group to perform undifferentiated labour. Furthermore, growth carries a coordination weight: larger organizations need to assign some of its members the role of coordinating the work of others, or of interfacing with other of its groups. That is, larger organizations have needs that are caused by their size, and that smaller organizations do not have. With organizations in the range we observed, some degree of specialization is well underway. We expect that, should we perform this study in a much larger organization, we would discover further differentiation between the roles of its members.

Growth-driven role specialization can be resisted and overturned, as some Agile proposals suggest. In this sense, however, the most interesting event in our observations was an attempt to ``break silos'' at Cobalt, which was successful for its stated goals, but \emph{increased} role differentiation in the organization. The silos that group members wanted to break consisted of division of labour by project. They discovered that assigning one or a few persons per project introduced vulnerabilities (when something went wrong in a project, only one or a few persons could intervene) and rework (as many projects consisted of similar modules that had to be developed independently). To break these silos, Cobalt assigned five of its senior developers the position of Subsystem Leads, so that each of them would be the first point of contact for matters concerning their subsystem. The end result was a greater differentiation between the roles of the five Subsystem Leads (previously simply Developers), but also a perceived decrease on the level of isolation, as projects begun to cross-pollinate and Subsystem Leads took care to help their teammates understand the particularities of their subsystems.

The other important factor driving role evolution in our findings was a mismatch between expectations and behaviour. The fact that expectations are placed does not mean that they are satisfied. There are many potential reasons for their lack of satisfaction: for instance, the person upon whom they are placed may not know or understand the expectation, or it could be beyond her means, or it could turn out to be in conflict with other expectations. Therefore, an assignment of expectations that work in theory may not work in practice. Similarly, an assignment that \emph{does not} work in theory may work in practice, due to the initiative or motivation of some people who contribute beyond what the rest expects of them. However, in both cases (under- and over-performance), over time organization members grow to expect the typical performance of others, and a new understanding (a new expectation, and hence, a new role) takes hold.

Several examples from our data above illustrate this phenomenon; we offer one more here. The coordinator of the Science group at Cobalt was brought in to do research with the data that Cobalt serviced, and in a ``consultative role.'' As time progressed, and without being contractually obliged nor pressured by his peers, he begun taking other responsibilities: providing feedback on how users would use some service features, being one of the main contacts for scientists outside the group (which lends more domain legitimacy to the organization), and helping to steer the group's direction and next steps. In many ways, his role begun to look more like that of the Gallium Product Manager we discussed above, without being named or being perceived as a Product Manager by anybody in the organization. As a result, he now plays an essential role in the development of the group's products and in the maintenance of its relations with its users, a role that the rest of the organization depends on for its good functioning.

\subsection{Roles and Communication Media (RQ6)}

@@

% I think it will be worth listing all the important ones. Perhaps more than a list? The key ideas here are that (a) these expectations *need* to be satisfied, or the performance of the organization will suffer, (b) some of them are conceptually closer to each other, and having them be carried out by the same person or persons should be more effective from a coordination perspective (see Proportionality, Maturity), and (c) understanding these patterns of responsibility pairing should yield efficiency gains more generalized.
